---
title: "Analysis"
author: "Mihailo Joksimovic"
date: "2/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(modelr)
library(randomForest)

read_csv('./globalterrorismdb_0718dist.csv') -> df

small_set <- df %>% 
  filter(iyear >= 2013, iday != 0, imonth != 0) %>% 
  mutate(
    imonth = as.factor(imonth), 
    iday = as.factor(iday),
    success = as.factor(success)
  ) %>% 
  mutate(
    date = make_date(
      iyear, 
      as.numeric(levels(imonth))[imonth], # Proper way of converting factor to number
      as.numeric(levels(iday))[iday]) # Proper way of converting factor to number
  ) %>% 
  select(eventid, date, everything())

```

## Basic info

This analysis file is split into two main parts. First part is concerned with Exploratory Data Analysis, where
we'll focus on exploring what we have inside the dataset, and the second part is concerned with predictions.

## European Data

We'll start our exploration by focusing on the data we have. Let's see what we have in this dataset.

```{r df}
ggplot(df, aes(x = iyear)) + geom_bar()
```

What we can see here is that data is being tracked from 1970s and there seems to be a single year that is missing.

Let's see what the missing year is:

```{r}
df %>% filter(iyear > 1990, iyear < 2000) %>% distinct(iyear)
```
What we can see is that, for some reason, data seems to be missing for 1993. That's ok, I guess we can live with that.

According to the official docs (<https://www.start.umd.edu/gtd/downloads/Codebook.pdf>), the best data has been collected starting from 2011. Therefore, for sake of ease and readability, let's focus on the period after 2011.

```{r}
small_set <- df %>% filter(iyear >= 2013)

# How many observations do we have?
count(small_set)
```

Let's see the # of attacks now and, for ease of analysis, let's facet it by Region

```{r}
ggplot(small_set, aes(iyear)) + geom_bar() + facet_wrap(~ region_txt)
```

That gives an interesting clue. So seems like Middle East, Nortf Africa and South Asia have higher number of attacks. We'll put a better focus on that later on.

Let's see a # of attacks per region.

```{r}
by_region <- small_set %>%
  group_by(region_txt) %>% 
  summarise(n = n())

ggplot(by_region) + 
  geom_bar(aes(x = reorder(region_txt, n), y = n), stat = "identity") + 
  coord_flip() +
  xlab("Count") +
  ylab("Region")
```

So we can confirm that previous observation makes sense really.

Let's see what are the most affected countries actually.

```{r}
by_country <- small_set %>%
  group_by(country_txt) %>% 
  summarise(n = n()) %>% 
  arrange(country_txt, desc(n)) %>% 
  top_n(10, n)

# Plot only top 10 countries ...
ggplot(by_country) + 
  geom_bar(aes(x = reorder(country_txt, n), y = n), stat = "identity") + 
  coord_flip() +
  xlab("Count") +
  ylab("Region")
```

So seems like Iraq is the country with most registered attacks.

Let's explore a bit more now. Which methods of attack are present?

```{r}
by_attacktype <- small_set %>% 
  group_by(attacktype1_txt) %>% 
  summarise(n = n())

ggplot(by_attacktype) + 
  geom_bar(aes(x = reorder(attacktype1_txt, n), y = n), stat = "identity") + 
  coord_flip() +
  xlab("Count") +
  ylab("Attack Type")
```

So seems like Bombing & Explosion are favorite attack methods. After that comes the Armed Assault and then we have Kindapping. Well, the first one seems to be interesting to explore later on.

What are the most used weapon types?

```{r}
by_weapontype <- small_set %>%
  mutate(label = str_trunc(weaptype1_txt, 20)) %>% 
  group_by(label) %>% 
  summarise(n = n())

ggplot(by_weapontype) + 
  geom_bar(aes(x = reorder(label, n), y = n), stat = "identity") + 
  coord_flip() +
  xlab("Count") +
  ylab("Weapon Type")
```

So explosive seems to be most commonly used means of attack.

